---
title: 'STAT394 Group Project: Group 11'
author: "Tom Tribe, Ken MacIver, Jundi Yang, Mei Huang"
date: "`r Sys.Date()`"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r, echo=FALSE, message=FALSE, warning = FALSE}
# load required packages
library(ggplot2)
library(GGally)


```


## Group Members

Tom Tribe 
- tom.tribe2016@gmail.com
- 0000-0002-5002-8066

Ken MacIver
- ken.maciver68@gmail.com
- 0000-0001-8999-4598

Jundi Yang
-  ivyli112358@gmail.com
- 0000-0003-0888-9564 

Mei Huang 
- huangmei139@gmail.com
- 0000-0003-2401-0679


# The Diamonds dataset

- Dataset: Diamonds
- 53940 rows (diamonds)
- ten variables
- very large dataset (approx 540,000 values)
- slow to process!

# The Variables
 \textcolor{red}{red font = categorical variable}
 
-   carat: the diamond's weight
-   \textcolor{red}{cut: a measure of quality}
-   \textcolor{red}{color: a measure of colour quality}
-   \textcolor{red}{clarity: a measure of clearness}
-   x: length in mm
-   y: width in mm 
-   z: depth in mm
-   depth: total depth percentage
-   table: width of top of diamond relative to widest point
-   price: the price of the diamond in US dollars 

(List adapted from list at kaggle.com).

# The Response Variable

'Price' seemed to us to be the obvious response variable. 

# Data Visualization (the dataset)

```{r, echo = FALSE}
diamonds <- read.csv("./diamonds.csv", encoding = "UTF-8")
diamonds$X <- NULL
diamonds[1:5,]
```

# Data Visualisation (pairs plot)

```{r}
diamonds$X <- NULL
diamonds_num <- subset(diamonds, select = c(carat,depth,table,price,x,y,z))
ggpairs(diamonds_num)
```

# Other things of interest

The EDA revealed the following:

- some variables not Normally distributed
- long right tails due to a few very expensive diamonds
- some zero values

# Next Steps

- Principal Component Analysis
- Find best predictor variable for price

