---
title: "Diamonds - Group 11"
author: "Tom Tribe, Ken MacIver, Jundi Yang, Mei Huang"
date: "2022-09-07"
output: 
 ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Group Members

Tom Tribe 
- tom.tribe2016@gmail.com
- 0000-0002-5002-8066

Ken MacIver
- ken.maciver68@gmail.com
- 0000-0001-8999-4598

Jundi Yang
-  ivyli112358@gmail.com
- 0000-0003-0888-9564 

Mei Huang 
- huangmei139@gmail.com
- 0000-0003-2401-0679

## Dataset

- We have chosen a data set called 'diamonds', which presents data on 53940 diamonds. 
- It was accessed it from kaggle.com. 
- There are ten variables: 3 categorical and 7 numerical 

- We choose this data set as it fit the criteria for this project and because it was simple to understand what each variable was a measure of. We are particularly interested in which variables are most predicitve of diamond price. 

## Variables

-   carat: the diamond's weight (numerical: 0.2 - 5.01)
-   cut: a measure of quality (categorical: Fair, Good, Very Good, Premium)
-   color: a measure of colour quality (categorical: J, which is poorest quality, to D, which is best)
-   clarity: a measure of clearness (categorical: from worst to best = I1, SI2, VS2, VS1, VVS2, IF)
-   x: length in mm (0 - 10.74)
-   y: width in mm (0 - 58.9) 
-   z: depth in mm (0 - 31.8)
-   depth: total depth percentage = z/mean(x,y) = 2\*z/(x+y) (43 - 79)
-   table: width of top of diamond relative to widest point
-   price: the price of the diamond in US dollars 

## Data Visualization

```{r, echo = FALSE}

diamonds <- read.csv("./diamonds.csv", encoding = "UTF-8")
diamonds[1:5,]
diamonds$X <- NULL

a <- ggplot(data=diamonds_melt, aes(x=metrics, y=value)) +
  
  geom_boxplot(aes(col=metrics), notch = TRUE) +
  scale_y_log10()+
  facet_grid(~ cut) +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
        legend.position = "none")

b <- ggplot(data=diamonds_melt, aes(x=metrics, y=value)) +
  
  geom_boxplot(aes(col=metrics), notch = TRUE) +
  scale_y_log10()+
  facet_grid(~ cut) +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
        legend.position = "none")

c <- ggplot(data=diamonds_melt, aes(x=metrics, y=value)) +
  
  geom_boxplot(aes(col=metrics), notch = TRUE) +
  scale_y_log10()+
  facet_grid(~ color) +
  theme(axis.text.x = element_text(size=7, angle=90, hjust=1),
        legend.position = "none")

ggarrange(a, b, c,+ rremove("x.text"), 
          labels = c("Cut", "Color", "Clarity"),
          ncol = 3, nrow = 1)
```

## Data Visualization

## Other things of interest

## Next Steps
