
---
title: "Diamonds - Group 11"
author: "Tom Tribe, Ken MacIver, Jundi Yang, Mei Huang"
date: "2022-09-07"
output: 
 ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(ggplot2)
require(GGally)
```

## Group Members

Tom Tribe 
- tom.tribe2016@gmail.com
- 0000-0002-5002-8066

Ken MacIver
- ken.maciver68@gmail.com
- 0000-0001-8999-4598

Jundi Yang
-  ivyli112358@gmail.com
- 0000-0003-0888-9564 

Mei Huang 
- huangmei139@gmail.com
- 0000-0003-2401-0679

## Dataset

- We have chosen a data set called 'diamonds', which presents data on 53940 diamonds. 
- It was accessed it from kaggle.com. 
- There are ten variables: 3 categorical and 7 numerical 

- We choose this data set as it fit the criteria for this project and because it was simple to understand what each variable was a measure of. We are particularly interested in which variables are most predicitve of diamond price. 

## Variables

<span style="color: red;">red font = categorical variable</span>

-   carat: the diamond's weight
-   <span style="color: red;">cut: a measure of quality</span>
-   <span style="color: red;">color: a measure of colour quality</span>
-   <span style="color: red;">clarity: a measure of clearness</span>
-   x: length in mm
-   y: width in mm 
-   z: depth in mm
-   depth: total depth percentage
-   table: width of top of diamond relative to widest point
-   price: the price of the diamond in US dollars 

(List adapted from list at kaggle.com).

## Data Visualization

```{r, echo = FALSE}
diamonds <- read.csv("./diamonds.csv", encoding = "UTF-8")
diamonds$X <- NULL
diamonds[1:5,]

```

## Data Visualization

```{r}

diamonds_num <- subset(diamonds, select = c(carat,depth,table,price,x,y,z))
ggpairs(diamonds_num)
```



## Other things of interest

The EDA revealed the following:

- some variables not Normally distributed
- long right tails due to a few very expensive diamonds
- some zero values


## Next Steps

- Principal Component Analysis
- Find best predictor variable for price