---
title: "Diamonds"
author: "Mei Huang, 300504502"
date: "`r Sys.Date()`"
header-includes:
    - \usepackage{newpxtext,eulerpx}
    - \usepackage{bm,bbm}
    - \usepackage{booktabs}

    
output: bookdown::pdf_document2

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed((300504502), kind="Mersenne-Twister")

library(ggplot2)
library(ggcorrplot)
library(ggthemes)
library(psych)
library(dplyr)
library(ggExtra)
library(moments)
library(fitdistrplus)
```

# Q2
From summary(Table&nbsp;\ref{tab:summary2}), we can see the diamonds dataset has 7 number variables and 3 category variables. We can extract numeric variables and have a look of their distribution.

From correlation visualization ((Figure&nbsp;\@ref(fig:correlation2))), pairs plot((Figure&nbsp;\@ref(fig:pairsplot2))) we can see there is very strong correlation among carat, x, y, z, and price.It seems obvious that the diamonds' weight, length, width, depth are all related to its size and size is the main predictor for diamonds price.  

The scatterplots and marginal boxplots(Figure&nbsp;\@ref(fig:scatter21), Figure&nbsp;\@ref(fig:scatter23), Figure&nbsp;\@ref(fig:scatter24), Figure&nbsp;\@ref(fig:scatter25), Figure&nbsp;\@ref(fig:scatter26), Figure&nbsp;\@ref(fig:scatter28), Figure&nbsp;\@ref(fig:scatter36), Figure&nbsp;\@ref(fig:scatter37), Figure&nbsp;\@ref(fig:scatter38)) also reveal their highly positive correlation and all of these data are skewed to the right. Meaning the large size diamonds are rear and the relationship among them may be different. The y z values are very concentrated

From Cullen and Frey plots, we can see the distribution of carat (Figure&nbsp;\@ref(fig:CF21)), price(Figure&nbsp;\@ref(fig:CF24)), x(Figure&nbsp;\@ref(fig:CF25)) are beta distribution, 
The distribution of depth (Figure&nbsp;\@ref(fig:CF22)), table(Figure&nbsp;\@ref(fig:CF23)), y(Figure&nbsp;\@ref(fig:CF26)), z(Figure&nbsp;\@ref(fig:CF27)) are not clear.

```{r}
df<-read.csv("diamonds.csv")
summary(df)

```

```{r}
# get a distribution of categorical variables.
count(df, cut)
count(df, color)
count(df, clarity)

```

```{r}
#choose numeric variables
df2<-df[, -c(1, 3, 4, 5)]
```


```{r, echo=FALSE}

```{r, echo=FALSE}
MySummary <- function(x){
  return(c(
    "minimum" = min(x),
    "first quartile" =quantile(x, .25),
    "median" =median(x),
    "third quartile" =quantile(x, .75),
    "maximum" = max(x),
    "IRQ" = IQR(x),
    "sd" = sd(x),
    "skewness" =skewness(x), 
    "kurtosis" = kurtosis(x))
  )
}

summary<-apply(df2, MySummary, MARGIN=2)
```

```{r, echo=FALSE}
knitr::kable(summary, digits=2, 
             caption="Statistic summary of diamonds", 
             label="summary2", 
             format="latex", booktabs=TRUE)
```

```{r}
#mean values for each eigenvalues
colMeans(df2)
#covariance matrix
round(var(df2), 3)
#Correlation Matrix
round(cor(df2), 3)
```
 
```{r correlation2, fig.cap="The visualtion of Diamonds correlation matrix", fig.pos="center", echo=FALSE}
 
 ggcorrplot(cor(df2),
           method = "circle", 
           hc.order = TRUE, 
           type = "lower")
```
 

```{r pairsplot2, fig.cap="The pairs plot for Diamonds numeric values", fig.pos="center", echo=FALSE}
#the pairs plot
pairs.panels(df2, method = "spearman", 
              #hist.col = "lightgreen", 
              density = TRUE, 
              ellipses = FALSE)
```


```{r scatter21, fig.cap="Scatterplot for carat & depth with its marginal boxplot", fig.pos="center", echo=FALSE}
p21<- ggplot(df2, aes(x = carat, y = depth))+
  geom_point()
p21<-ggMarginal(p21, type = "boxplot", notch = TRUE, size =7)
p21
   
```

```{r scatter22, fig.cap="Scatterplot for carat & table with its marginal boxplot", fig.pos="center", echo=FALSE}
p22<- ggplot(df2, aes(x = carat, y = table))+
  geom_point()
p22<-ggMarginal(p22, type = "boxplot", notch = TRUE, size =7)
p22
   
```

```{r scatter23, fig.cap="Scatterplot for carat & price with its marginal boxplot", fig.pos="center", echo=FALSE}
p23<- ggplot(df2, aes(x = carat, y = price))+
  geom_point()
p23<-ggMarginal(p23, type = "boxplot", notch = TRUE, size =7)
p23
   
```

```{r scatter24, fig.cap="Scatterplot for carat & x with its marginal boxplot", fig.pos="center", echo=FALSE}
p24<- ggplot(df2, aes(x = carat, y = x))+
  geom_point()
p24<-ggMarginal(p24, type = "boxplot", notch = TRUE, size =7)
p24
```

```{r scatter25, fig.cap="Scatterplot for carat & y with its marginal boxplot", fig.pos="center", echo=FALSE}
p25<- ggplot(df2, aes(x = carat, y = y))+
  geom_point()
p25<-ggMarginal(p25, type = "boxplot", notch = TRUE, size =7)
p25
   
```

```{r scatter26, fig.cap="Scatterplot for carat & z with its marginal boxplot", fig.pos="center", echo=FALSE}
p26<- ggplot(df2, aes(x = carat, y = z))+
  geom_point()
p26<-ggMarginal(p26, type = "boxplot", notch = TRUE, size =7)
p26
   
```

```{r scatter27, fig.cap="Scatterplot for depth & table with its marginal boxplot", fig.pos="center", echo=FALSE}
p27<- ggplot(df2, aes(x = depth, y = table))+
  geom_point()
p27<-ggMarginal(p27, type = "boxplot", notch = TRUE, size =7)
p27
   
```

```{r scatter28, fig.cap="Scatterplot for depth & price with its marginal boxplot", fig.pos="center", echo=FALSE}
p28<- ggplot(df2, aes(x = depth, y = price))+
  geom_point()
p28<-ggMarginal(p28, type = "boxplot", notch = TRUE, size =7)
p28
   
```

```{r scatter29, fig.cap="Scatterplot for depth & x with its marginal boxplot", fig.pos="center", echo=FALSE}
p29<- ggplot(df2, aes(x = depth, y = x))+
  geom_point()
p29<-ggMarginal(p29, type = "boxplot", notch = TRUE, size =7)
p29
   
```

```{r scatter30, fig.cap="Scatterplot for depth & y with its marginal boxplot", fig.pos="center", echo=FALSE}
p30<- ggplot(df2, aes(x = depth, y = y))+
  geom_point()
p30<-ggMarginal(p30, type = "boxplot", notch = TRUE, size =7)
p30
   
```

```{r scatter31, fig.cap="Scatterplot for depth & z with its marginal boxplot", fig.pos="center", echo=FALSE}
p31<- ggplot(df2, aes(x = depth, y = z))+
  geom_point()
p31<-ggMarginal(p31, type = "boxplot", notch = TRUE, size =7)
p31
   
```

```{r scatter32, fig.cap="Scatterplot for table & price with its marginal boxplot", fig.pos="center", echo=FALSE}
p32<- ggplot(df2, aes(x = table, y = price))+
  geom_point()
p32<-ggMarginal(p32, type = "boxplot", notch = TRUE, size =7)
p32
   
```

```{r scatter33, fig.cap="Scatterplot for table & x with its marginal boxplot", fig.pos="center", echo=FALSE}
p33<- ggplot(df2, aes(x = table, y = x))+
  geom_point()
p33<-ggMarginal(p33, type = "boxplot", notch = TRUE, size =7)
p33
   
```

```{r scatter34, fig.cap="Scatterplot for table & y with its marginal boxplot", fig.pos="center", echo=FALSE}
p34<- ggplot(df2, aes(x = table, y = y))+
  geom_point()
p34<-ggMarginal(p34, type = "boxplot", notch = TRUE, size =7)
p34
   
```

```{r scatter35, fig.cap="Scatterplot for table & z with its marginal boxplot", fig.pos="center", echo=FALSE}
p35<- ggplot(df2, aes(x = table, y = price))+
  geom_point()
p35<-ggMarginal(p35, type = "boxplot", notch = TRUE, size =7)
p35
   
```

```{r scatter36, fig.cap="Scatterplot for price & x with its marginal boxplot", fig.pos="center", echo=FALSE}
p36<- ggplot(df2, aes(x = price, y = x))+
  geom_point()
p36<-ggMarginal(p36, type = "boxplot", notch = TRUE, size =7)
p36
   
```

```{r scatter37, fig.cap="Scatterplot for price & y with its marginal boxplot", fig.pos="center", echo=FALSE}
p37<- ggplot(df2, aes(x = price, y = y))+
  geom_point()
p37<-ggMarginal(p37, type = "boxplot", notch = TRUE, size =7)
p37
   
```


```{r scatter38, fig.cap="Scatterplot for price & z with its marginal boxplot", fig.pos="center", echo=FALSE}
p38<- ggplot(df2, aes(x = price, y = z))+
  geom_point()
p38<-ggMarginal(p38, type = "boxplot", notch = TRUE, size =7)
p38
   
```



```{r CF21, fig.cap="Cullen & Frey plot for carat", fig.pos="center", echo=FALSE}

descdist(df2$carat, discrete=FALSE, boot=500)
```

```{r CF22, fig.cap="Cullen & Frey plot for depth", fig.pos="center", echo=FALSE}
descdist(df2$depth, discrete=FALSE, boot=500)
```

```{r CF23, fig.cap="Cullen & Frey plot for table", fig.pos="center", echo=FALSE}
descdist(df2$table, discrete=FALSE, boot=500)
```

```{r CF24, fig.cap="Cullen & Frey plot for price", fig.pos="center", echo=FALSE}
descdist(df2$price, discrete=FALSE, boot=500)
```
```{r CF25, fig.cap="Cullen & Frey plot for x", fig.pos="center", echo=FALSE}
descdist(df2$x, discrete=FALSE, boot=500)
```

```{r CF26, fig.cap="Cullen & Frey plot for y", fig.pos="center", echo=FALSE}
descdist(df2$y, discrete=FALSE, boot=500)
```

```{r CF27, fig.cap="Cullen & Frey plot for z", fig.pos="center", echo=FALSE}
descdist(df2$z, discrete=FALSE, boot=500)
```

# Reference
SHIVAM AGRAWAL.(2017) "Diaminds"
URL:https://www.kaggle.com/datasets/shivam2503/diamonds